Author: Olivia Holt

note: from eds 223 course in the MEDS program at UCSB

# Overview

Phenology is the timing of life history events. Important phenological events for plants involve the growth of leaves, flowering, and senescence (death of leaves). Plants species adapt the timing of these events to local climate conditions to ensure successful reproduction. Subsequently, animal species often adapt their phenology to take advantage of food availability. As the climate shifts this synchronization is being thrown out of whack. Shifts in phenology are therefore a common yardstick of understanding how and if ecosystems are adjusting to climate change.\

Plant species may employ the following phenological strategies:\

-   winter deciduous: lose leaves in the winter, grow new leaves in the spring\
-   drought deciduous: lose leaves in the summer when water is limited\
-   evergreen: maintain leaves yearround\

credit: this lab is based on a materials developed by Chris Kibler.

# Tasks

To analyze plant phenology near the Santa Clara River which flows from Santa Clarita to Ventura and investigate the phenology of the following plant communities:\

-   riparian forests: grow along the river, dominated by winter deciduous cottonwood and willow trees\
-   grasslands: grow in open spaces, dominated by drought deciduous grasses\
-   chaparral shrub lands: grow in more arid habitats, dominated by evergreen shrubs\

A time series of Landsat imagery and polygons identifying the locations of study sites within each plant community will be used to investigate the phenology of the specified plant communities.\

## Data

**Landsat Operational Land Imager (OLI sensor)**\

-   8 pre-processed scenes\
    -   Level 2 surface reflectance products\
    -   erroneous values were set to NA\
    -   scale factor set to 100\
    -   bands 2-7\
    -   dates in filenname\

**Study sites**

-   polygons representing sites\
    -   study_site: character string with plant type

## Summary of approach

-   Convert spectral relfectance into a measure of vegetation productivity (NDVI)
-   Calulate NDVI throughout the year
-   Summarize NDVI values within vegetation communities
-   Visualize changes in NDVI within vegetation communities

# Workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
```

```{r include=TRUE, message=FALSE}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(here)
library(tmap)
library(cowplot)

rm(list = ls())
#here::i_am("labs/week8.Rmd")
```

### create an NDVI function

Defining a function to compute the Normalized Difference Vegetation Index (NDVI). NDVI computes the difference in reflectance in the near infrared and red bands, normalized by their sum.

```{r ndvi-function}
ndvi_fun <- function(nir, red){
  (nir - red)(nir + red)
}

```

### computing NDVI for a single scene

Starting with the first scene collected on June 12, 2018

```{r}
#reading in a raster
landsat_20180612 <- rast(here("MEDS","Course","eds223-geospatial","labs","week-8","data","landsat_20180612.tif"))
```

Update the names of the layers to match the spectral bands they correspond to

```{r}
#updating the names layer
names(landsat_20180612) <- c("blue","green","red","NIR","SWIR1","SWIR2")
```

Apply the NDVI function to compute NDVI for this scene using the `lapp()` function with the NIR and red layers as arguments.

```{r}
ndvi_20180612 <- lapp(landsat_20180612[[c("NIR","red")]], fun = ndvi_fun)
```

### computing NDVI for all scenes

Using the function from above for all scenes to minimize copy/pasting.

Clear the environment and redefine the function for NDVI.

```{r}
rm(list = ls())
#here::i_am("labs/week8.Rmd")

ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
```

#### outlining the function

sketching out a workflow for the function

```{r eval=FALSE}

# note: this code is not meant to run!

# create_ndvi_layer <- function(file) {
#   landsat <- rast(file)
#   names(landsat) <- c("blue","green","red","NIR","SWIR1","SWIR2")
#   ndvi <- lapp(landsat[[c(4,3)]], ndvi_fun)
#   return(ndvi)
# }




```

#### creating a list of files/scenes

Create a list of the scenes for the function to compute NDVI for each.

```{r}
# here we're asking for the names of all the files in the week8 folder
# the "pattern" option returns names that end in .tif (the file extension for the landsat scences)
# the "full.names" option returns the full file path for each scene

list.files(here("MEDS","Course","eds223-geospatial","labs","week-8","data"), pattern = "*.tif", full.names = TRUE)
#you could put "*.tif$" then it has to end in ".tif"


```

#### updating our function

Now let's update our function to work with list of file names we created

```{r}

# now we're passing our function a number that will correspond to the index in the list of file names

create_ndvi_layer <- function(i) {
  file <- files[i]
  landsat <- rast(file[i])
  names(landsat) <- c("blue","green","red","NIR","SWIR1","SWIR2")
  ndvi <- lapp(landsat[[c(4,3)]], ndvi_fun)
  return(ndvi)
 }

# let's test our function by asking it to read in the first file 

test <- create_ndvi_layer(1)

```

#### run function on all scenes!

Now we can use our function to create a NDVI layer for each scene and stack them into a single rasterstack.

```{r}
# create NDVI layer for each scene (day) and stack into a single rasterstack

all_ndvi <- c(create_ndvi_layer(1),
              create_ndvi_layer(2),
              create_ndvi_layer(3),
              create_ndvi_layer(4),
              create_ndvi_layer(5),
              create_ndvi_layer(6),
              create_ndvi_layer(7),
              create_ndvi_layer(8))

# update layer names to match date

names(all_ndvi) <- c("2018-06-12",
                     "2018-08-15",
                     "2018-10-18",
                     "2018-11-03",
                     "2019-01-22",
                     "2019-02-23",
                     "2019-04-12",
                     "2019-07-01")

```

### Compare NDVI across vegetation communities

Now that we have computed NDVI for each of our scenes (days) we want to compare changes in NDVI values across different vegetation communities.

#### read in study sites

First, we'll read in a shapefile of study sites.

```{r include=TRUE}

sites <- st_read(here("MEDS","Course","eds223-geospatial","labs","week-8","data","study_sites.shp"))

# plot study sites on a single NDVI layer

tm_shape(all_ndvi[[1]])+
  tm_raster()+
  tm_shape(sites)+
  tm_polygons()

```

#### extract NDVI at study sites

Here we find the average NDVI within each study site. The output of extract is a data frame with rows that match the study site dataset, so we bind the results to the original dataset.

```{r include=TRUE}

sites_ndvi <- terra::extract(all_ndvi, sites, fun = "mean")

sites_annotated <- cbind(sites, sites_ndvi)
```

#### clean results

We're done! Except our data is very untidy... Let's tidy it up!\

-   convert to data frame
-   turn from wide to long format
-   turn layer names into date format

```{r include=TRUE}

sites_clean <- sites_annotated %>% 
  st_drop_geometry() %>% 
  select(-ID) %>% 
  pivot_longer(!study_site) %>% 
  rename(NDVI = value) %>% 
  mutate("year" = str_sub(name,2,5),
         "month" = str_sub(name,7,8),
         "day" = str_sub(name,-2,-1)) %>% 
  unite("date", 4:6, sep = "-") %>% 
  mutate("date" = lubridate::as_date(date)) %>% 
  select(-name)

```

#### plot results

```{r include=TRUE}

ggplot(sites_clean,
       aes(x = date, y = NDVI,
           group = study_site,
           col = study_site))+
  geom_line()+
  theme_minimal()+
  labs(x = "", y = "Normalized Difference Vegetation Index (NDVI)",
       col = "Vegetation type",
       title =  "Seasonal cycles of vegetation productivity")

```
